### 2021-11-08 chapter2/chapter.tex
first steps of nigels changes to first part of ch 3. up to sec 3.3. about to change to matrix notation instead of partition into phases.

Here, the construction of the approximating QBD-RAP is developed intuitively, from a stochastic modelling perspective. The key observation is that, given \(\{\varphi(t)\}\) is constant, then \(\{X(t)\}\) moves deterministically. Upon discretising the state space of the level process into intervals of width \(\Delta\), then, given \(X(t)\), the distribution of time it takes for \(\{X(t)\}\) to leave a given interval on the event that \(\{\varphi(t)\}\) remains in the same phase, is deterministic. We model this deterministic behaviour approximately by concentrated matrix exponential distributions (matrix exponential distributions with low-variance) \cite{ert2006,hstz2016,ert2006,hht2020,mt2021}. We construct a level process for the QBD-RAP and correspond it to a discretisation the level process of the fluid queue. This is similar to \cite{bo2013} where the level process of their QBD corresponds a discretisation of the fluid level. Unlike \cite{bo2013}, however, here we do not take \(\Delta \to 0\). Rather, we suppose that the variance of the concentrated matrix exponential distributions we use to approximate deterministic events gets small. % This necessarily means that the order of the matrix exponential increases.

@@

We interpret the position of the orbit \(\bs A(t+u) = \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}\) as corresponding to the fact that \(X(t+u)\) is \(|c_i|u\) units from the left-hand boundary of the interval \(\calD_{k,i}\). This gives a heuristic argument as to how we can model the sojourn times in a given interval \(\mathcal D_{k,i}\) on the event that the phase does not change.

We can apply analogous arguments to heuristically develop a model for the sojourn time of the fluid queue in an interval, \(\calD_{k,i}\), \(i\in\calS_-\), on the event that the phase does not change, given the fluid is \(X(t)=y_{k+1}\). For \(i\in\calS_-\), the orbit position \(\bs A(t+u) = \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}\) is interpreted as corresponding to \(X(t)\) being \(|c_i|u\) units from the right-hand boundary of the interval \(\calD_{k,i}\), \(y_{k+1}\).

@@

Let \(E\) be the event that, at time \(t\) the orbit of the QBD-RAP is \(\bs A(t) = \bs \alpha\), the phase process is \(\varphi(t) = i\in\calS_+\), and there are no events of the QBD-RAP before time \(t+u\). On the event \(E\) the orbit position at time \(t+u\) will be 
\begin{align}
	\bs A(t+u) = \cfrac{\bs \alpha e^{\bs S_i u}}{\bs \alpha e^{\bs S_i u}\bs e}.\label{eqn: orbit i}
\end{align}
Correspondingly, on the event that at time \(t\) the level process of the fluid queue is \(X(t)=y_{k}\), the phase is \(\varphi(t)=i\) and there are no change of phase by time \(t+u\), then \(X(t+u)=y_{k}+c_iu\). That is, \(X(t+u)\) is \(c_iu\) units from the left-hand edge of \(\calD_k\).

If, at time \(t+u\) there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_+\), then we need to map the orbit position \(\bs A(t+u^-) = \cfrac{\bs \alpha e^{\bs S_i u}}{\bs \alpha e^{\bs S_i u}\bs e}\) to an orbit position which corresponds to the being \(c_iu\) units from the left-hand edge of \(\calD_k\) in phase \(j\). Call this mapping \(\bs D(i,j)(\cdot)\). As noted in Section \ref{sec: qbd-rap}, the mapping must be linear, 
\(\bs D(i,j)(\bs a) = \bs a \bs D(i,j),\, \bs a \in\mathcal A,\)
for some matrix \(\bs D(i,j)\). The matrix \(\bs D(i,j)\) is a modelling choice.

@@

An orbit position which corresponds to the level process of the QBD-RAP beign \(c_iu\) units to the right of \(y_k\) in phase \(j\in\calS_+\) is \(\bs A(t+u) = \cfrac{\bs \alpha e^{\bs S_j (|c_i|u)/|c_j|}}{\bs \alpha e^{\bs S_j (|c_i|u)/|c_j| u}\bs e}\). To see this, consider the event that at time \(t\) the orbit process of the QBD-RAP is \(\bs A(t) = \bs \alpha\), the phase process is \(\varphi(t) = j\in\calS_+\), and there are no events of the QBD-RAP before time \(t+|c_i|u/|c_j|\). On this event, the orbit position at time \(t+u\) will be \(\displaystyle \bs A(t+u) = \cfrac{\bs \alpha e^{\bs S_j (|c_i|u)/|c_j|}}{\bs \alpha e^{\bs S_j (|c_i|u)/|c_j| u}\bs e}\). Correspondingly, on the event that at time \(t\) the level process of the fluid queue is \(X(t)=y_{k}\), the phase is \(\varphi(t)=j\) and there are no change of phase by time \(t+|c_i|u/|c_j|\), then \(X(t+u)=y_{k}+c_j|c_i|u/|c_j| = y_k+c_iu\). That is, \(X(t+u)\) is \(c_iu\) units from the left-hand edge of \(\calD_k\).

Now, observe 
\[\cfrac{\bs \alpha e^{\bs S_j (|c_i|u)/|c_j|}}{\bs \alpha e^{\bs S_j (|c_i|u)/|c_j| u}\bs e} = \cfrac{\bs \alpha e^{\bs S_iu}}{\bs \alpha e^{\bs S_i u}\bs e},\] 
which is exactly (\ref{eqn: orbit i}), the orbit position on the event \(E\). Hence, it is sensible to choose \(\bs D(i,j)=\bs I\).

@@

\subsection{On a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\)} 
Now suppose \(X(t)=y_k\), \(\varphi(t)=i\in\calS_+\), and the phase remains in state \(i\) until there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), \(u\in[0,\Delta/|c_i|)\). We have reasoned, intuitively, that, for \(i\in\calS_+\) and for \(x=|c_i|u\in[0,\Delta)\), the orbit position \(\cfrac{\bs\alpha e^{\bs Sx} }{\bs\alpha e^{\bs Sx } \bs e}\) corresponds to the fluid level position \(X(t) = y_k+x\), that is \(x\) units greater than \(y_k\).

As before, we need to find a matrix \(\bs D(i,j)\) to map the orbit position on the event \(E\) when there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), from \(\bs A(t+u^-) = \cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\), to 
\[\bs A(t+u) = \cfrac{\bs A(t+u^-)\bs D(i,j)}{\bs A(t+u^-)\bs D(i,j)\bs e}=\cfrac{\cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\bs D(i,j)}{\cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\bs D(i,j)\bs e}=\cfrac{\bs\alpha e^{\bs S_iu} \bs D(i,j)}{\bs\alpha e^{\bs S_iu}\bs D(i,j)\bs e}\]
in such a way that the orbit position after the jump corresponds, in some way, to the fluid level being a distance of \(|c_i|u\) from \(y_k\) in phase \(j\).

@@

In the discussions above, we have relied on the relationship between the event \(E\) and the orbit position \(\displaystyle \bs A(t+u) = \cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\). The relationship allows us to associate this orbit position with the level of the fluid queue. For phase-type distributions the vector \(\cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\) is the vector of posterior probabilities \(\left[\mathbb P(J(u)=k \mid E)\right]_{k\in\{1,...,p\}}\). We can use this vector of posterior probabilities in the same way as we do for matrix-exponential distributions and QBD-RAPs. However, the orbit interpretation forgoes the presence of the phase process \(\{J(t)\}\). Here, we will use the phase process \(\{J(t)\}\) to derive a choice of \(\bs D(i,j)\).

Recall the position of the orbit process can be used to determine the residual time. This idea can be replicated when we are given the phase of the phase-type rather than the value of the orbit process. Given the phase of the phase-type distribution is \(k\in\{1,...,p\}\), the distribution of the residual time is 
\[\mathbb P(R_i\leq r \mid \mbox{phase}=k)=1-\bs e_k e^{\bs S_i r}\bs e.\]
Notice that this is independent of the time since the phase-type distribution was initialised. We call this time since the phase-type distribution was initialised the \emph{age}. As noted by \cite{hmp2017}, the distribution of the age given the phase of the phase-type is \(k\in\{1,...,p\}\) depends on the sampling scheme which determines the time of the observation. Here, on the event that at time \(t\) the phase of the QBD-RAP is \(i\), a change of phase occurs after an exponential amount of time with rate \(-T_{ii}\). Proposition 4.1, \cite{hmp2017} states that the distribution function of the age, given the phase of the phase-type is \(k\) and the process is observed after an exponential time with rate \(-T_{ii}\) is 
\[\mathbb P(\mbox{age}\leq u \mid \mbox{phase}=k )=1-\cfrac{\bs \alpha e^{(\bs S_i + T_{ii}\bs I)u}(-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e_k}{\bs \alpha (-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e_k}.\]
Let \(\widehat{ \bs S}_i(T_{ii}) = diag(\bs \nu)^{-1}\bs S_i'diag(\bs \nu)\), where \(\bs \nu = \bs\alpha(-(\bs S_i+ T_{ii}\bs I))^{-1}/(\bs\alpha(-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e)\). Algebraic manipulations show
\begin{align}
	1-\cfrac{\bs \alpha e^{(\bs S_i + T_{ii}\bs I)u}(-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e_k}{\bs \alpha (-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e_k}
	= 1-\bs e_k' e^{(\widehat{\bs S}_i(T_{ii}) + T_{ii}\bs I)u} \bs e,\label{eqn: skgj}
\end{align}
which is of Phase-type.

Recall that in the matrix-exponential case we associated the value of the orbit process \(\cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\) with being \(c_iu\) units to the right of \(y_{k}\). In contrast, for the phase-type case if we use the phase process \(\{J(t)\}\) to inform our modelling of the fluid queue (as opposed to the value of the orbit process), then we associate phase \(k\in\{1,...,p\}\) of \(\{J(t)\}\) with being a \emph{random} distance to the right of \(y_k\) where this distance has distribution 
\begin{align}
	1-\bs e_k' e^{(\widehat{\bs S}_i(T_{ii}) + T_{ii}\bs I)u} \bs e.\label{eqn: skgj2}
\end{align}
Therefore, on the event \(E\), upon a change from of phase of the QBD-RAP from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), it seems reasonable to want the distribution of time until the next event of the QBD-RAP to be
\begin{align}
	1-\bs e_k' e^{(\widehat{\bs S}_i(T_{ii}) + T_{ii}\bs I)|c_j|u/|c_i| + T_{jj}Iu} \bs e.\label{eqn: skgj3}
\end{align}
The factor \(|c_j|u/|c_i|\) arises as a conversion between the speed at which the fluid level moves in phase \(j\) compared to phase \(i\). The rate \(T_{jj}\) in the exponent is the rate at which the next change of phase occurs.

@@

For matrix exponential distributions we cannot rely on the phase process \(\{J(t)\}\) as we did in the phase-type case. Recall that we want to find a matrix \(\bs D(i,j)\) to map the orbit position on the event \(E\) when there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), from \(\bs A(t+u^-) = \cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\), to 
\[\bs A(t+u) = \cfrac{\bs A(t+u^-)\bs D(i,j)}{\bs A(t+u^-)\bs D(i,j)\bs e}=\cfrac{\cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\bs D(i,j)}{\cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\bs D(i,j)\bs e}=\cfrac{\bs\alpha e^{\bs S_iu} \bs D(i,j)}{\bs\alpha e^{\bs S_iu}\bs D(i,j)\bs e}\]
in such a way that the orbit position after the jump corresponds, in some way, to the fluid level being a distance of \(|c_i|u\) from \(y_k\) in phase \(j\).

Given our interpretation of the orbit position, \(\cfrac{\bs\alpha e^{\bs Sx} }{\bs\alpha e^{\bs Sx } \bs e}\), a solution would be find a mapping which takes \(\cfrac{\bs\alpha e^{\bs Sx} }{\bs\alpha e^{\bs Sx } \bs e}\) and maps it directly to \(\cfrac{\bs\alpha e^{\bs S(\Delta-x)} }{\bs\alpha e^{\bs S(\Delta-x) } \bs e}\). However, we have been unsuccessful in finding such a mapping. The main hurdle being that the map must be linear. Instead, we approximate as follows.

Recall that we use \(R_i(u)\) to approximate the distance of the fluid level to the left of \(y_{k+1}\). Suppose we are given \(R_i(u)=r\) which corresponds, approximately, to the fluid level being \(|c_i|r\) units to the left of \(y_{k+1}\). The position of the orbit process which corresponds to a distance of \(|c_i|r\) to the left of \(y_{k+1}\) in phase \(j\) is \(\cfrac{\bs\alpha e^{ \bs S_j|c_i|r/|c_j| } }{\bs\alpha e^{\bs S_j |c_i|r/|c_j| } \bs e}=\cfrac{\bs\alpha e^{ \bs S_ir } }{\bs\alpha e^{\bs S_ir } \bs e}\). Hence, on \(R_i(u)=r\), the event \(E\) and on the event that a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) occurs at time \(t+u\), the orbit position should jump from \(\cfrac{\bs\alpha e^{\bs S_iu} }{\bs\alpha e^{\bs S_iu } \bs e}\) to \(\cfrac{\bs\alpha e^{ \bs S_ir } }{\bs\alpha e^{\bs S_ir } \bs e}\). However, at time \(t+u\), \(R_i(u)\) is a random variable about the future of the process and therefore not known, so instead, we take the expected initial vector 
\[\mathbb E \left[ \cfrac{\bs\alpha e^{ \bs S_iR_i(u) } }{\bs\alpha e^{\bs S_iR_i(u) } \bs e} \right] = \int_{r=0}^\infty \cfrac{\bs\alpha e^{\bs S_i u } }{\bs\alpha e^{\bs S_iu } \bs e} e^{\bs S_ir } \bs s_i \cfrac{\bs\alpha e^{ \bs S_ir } }{\bs\alpha e^{ \bs S_ir } \bs e}\wrt r = \cfrac{\bs\alpha e^{\bs S_i u } }{\bs\alpha e^{\bs S_iu } \bs e} \int_{r=0}^\infty e^{\bs S_ir } \bs s_i\cfrac{\bs\alpha e^{ \bs S_ir } }{\bs\alpha e^{ \bs S_ir } \bs e}\wrt r.\]
After a change of variables \(x=|c_i|r\) we get 
\[\cfrac{\bs\alpha e^{\bs S_i u } }{\bs\alpha e^{\bs S_iu } \bs e} \int_{x=0}^\infty e^{\bs S x }|c_i| \bs s \cfrac{\bs\alpha e^{ \bs Sx } }{\bs\alpha e^{ \bs Sx } \bs e}\wrt x/|c_i| = \bs A(t+u^-)\int_{x=0}^\infty e^{\bs S x }\bs s \cfrac{\bs\alpha e^{ \bs Sx } }{\bs\alpha e^{ \bs Sx } \bs e}\wrt x,\]
since at time \(t+u^-\), the orbit position is \(\bs A(t+u^-) = \cfrac{\bs\alpha e^{\bs S_i u } }{\bs\alpha e^{\bs S_iu } \bs e}\). Thus, we find 
\[\bs D(i,j)=\displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \cfrac{\bs\alpha e^{ \bs Sx } }{\bs\alpha e^{ \bs Sx } \bs e}\wrt x=:\bs D.\]

Observe that \(\bs D\bs e = \displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \cfrac{\bs\alpha e^{ \bs Sx } }{\bs\alpha e^{ \bs Sx } \bs e}\wrt x\bs e =\displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \wrt x=\bs e\). Further, since \(\mathcal A\) is closed and convex \citep{MEinAP}, then \((\bs a \bs D,\bs S,\bs s)\) is a representation of a matrix exponential distribution for any \(\bs a\in\mathcal A\).

We pose the choice of the matrix \(\bs D\) as a modelling choice. Other choices are possible, for example, \(\displaystyle \bs D = \int_{x=0}^{\Delta-\varepsilon}e^{\bs S x }\bs s \cfrac{\bs\alpha e^{ \bs Sx } }{\bs\alpha e^{ \bs Sx } \bs e}\wrt x\). It may also be possible to construct other matrices \(\bs D\), perhaps via geometric arguments.

@@

At time \(t\) an event occurs at rate 
\[\bs A(t) (\bs{S}_i - T_{ii}\bs{I}_p)\bs e = \bs A(t)\bs s_i - T_{ii}.\]
More precisely, an event corresponding to a change in phase for \({\phi}(t)\) occurs at rate \(-\bs A(t) T_{ii}\bs e = -T_{ii}\) and an event corresponding to a \emph{change of level} occurs at rate \(-\bs A(t) \bs{S}_i\bs e = \bs A(t) \bs s_i\). Later, we will make clear why we say that the latter event corresponds to a change of level. Upon an event occurring at time \(t\), with probability \(\cfrac{-T_{ii}}{-T_{ii} + \bs A({t^-})\bs s_i}\) the event corresponds to a change of phase and with probability \(\cfrac{\bs A({t^-})\bs s_i}{-T_{ii} + \bs A({t^-})\bs s_i}\) the event corresponds a change of level.

>>>>

Here, the construction of the approximating QBD-RAP is developed intuitively, from a stochastic modelling perspective. The key observation is that, given \(\{\varphi(t)\}\) is constant, then \(\{X(t)\}\) moves deterministically. Upon discretising the state space of the level process into intervals of width \(\Delta\), then, given \(X(t)\), the distribution of time it takes for \(\{X(t)\}\) to leave a given interval on the event that \(\{\varphi(t)\}\) remains in the same phase, is deterministic. We model this deterministic behaviour approximately by concentrated matrix exponential distributions (matrix exponential distributions with low-variance) \citep{ert2006,hstz2016,ert2006,hht2020,mt2021}. We construct a level process for the QBD-RAP and correspond it to a discretisation the level process of the fluid queue. This is similar to \cite{bo2013} where the level process of their QBD corresponds a discretisation of the fluid level. Unlike \cite{bo2013}, however, here we do not take \(\Delta \to 0\). Rather, we suppose that the variance of the concentrated matrix exponential distributions we use to approximate deterministic events gets small. % This necessarily means that the order of the matrix exponential increases.

@@

For notational convenience, define the row-vector-valued function \(\bs k(t):\mathbb R\to \mathcal A\), 
\begin{equation}
	\bs k(t) = \cfrac{\bs \alpha e^{\bs S t}}{\bs \alpha e^{\bs S t}\bs e}, \label{eqn: orbit fun def}
\end{equation}
and the operator \(\bs K(t): \mathcal A\to \mathbb R\), \(t\geq 0\), 
\begin{equation}
	\bs a \bs K(t) = \cfrac{\bs a e^{\bs S t}}{\bs a e^{\bs S t}\bs e}, \label{eqn: operator fun def}
\end{equation}
for any \(a\in\mathcal A\).

We interpret the position of the orbit \(\bs A(t+u) = \bs k(|c_i|u)\) as corresponding to the fact that \(X(t+u)\) is \(|c_i|u\) units from the left-hand boundary of the interval \(\calD_{k,i}\). This gives a heuristic argument as to how we can model the sojourn times in a given interval \(\mathcal D_{k,i}\) on the event that the phase does not change.

We can apply analogous arguments to heuristically develop a model for the sojourn time of the fluid queue in an interval, \(\calD_{k,i}\), \(i\in\calS_-\), on the event that the phase does not change, given the fluid is \(X(t)=y_{k+1}\). For \(i\in\calS_-\), the orbit position \(\bs A(t+u) = \bs k(|c_i|u)\) is interpreted as corresponding to \(X(t)\) being \(|c_i|u\) units from the right-hand boundary of the interval \(\calD_{k,i}\), \(y_{k+1}\).

@@

Let \(E\) be the event that, at time \(t\) the orbit of the QBD-RAP is \(\bs A(t) = \bs \alpha\), the phase process is \(\varphi(t) = i\in\calS_+\), and there are no events of the QBD-RAP before time \(t+u\). On the event \(E\) the orbit position at time \(t+u\) will be 
\begin{align}
	\bs A(t+u) = \bs k(|c_i|u) = \cfrac{\bs \alpha e^{\bs S_i u}}{\bs \alpha e^{\bs S_i u}\bs e}.\label{eqn: orbit i}
\end{align}
Correspondingly, on the event that at time \(t\) the level process of the fluid queue is \(X(t)=y_{k}\), the phase is \(\varphi(t)=i\) and there are no change of phase by time \(t+u\), then \(X(t+u)=y_{k}+c_iu\). That is, \(X(t+u)\) is \(c_iu\) units from the left-hand edge of \(\calD_k\).

If, at time \(t+u\) there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_+\), then we need to map the orbit position \(\bs A(t+u^-) = \bs k(|c_i|u)\) to an orbit position which corresponds to the being \(c_iu\) units from the left-hand edge of \(\calD_k\) in phase \(j\). Call this mapping \(\bs D(i,j)(\cdot)\). As noted in Section \ref{sec: qbd-rap}, the mapping must be linear, 
\(\bs D(i,j)(\bs a) = \bs a \bs D(i,j),\, \bs a \in\mathcal A,\)
for some matrix \(\bs D(i,j)\). The matrix \(\bs D(i,j)\) is a modelling choice.

@@

An orbit position which corresponds to the level process of the QBD-RAP beign \(c_iu\) units to the right of \(y_k\) in phase \(j\in\calS_+\) is 
\[\bs A(t+u) = \cfrac{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|)}}{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|) u}\bs e}.\] 
To see this, consider the event that at time \(t\) the orbit process of the QBD-RAP is \(\bs A(t) = \bs \alpha\), the phase process is \(\varphi(t) = j\in\calS_+\), and there are no events of the QBD-RAP before time \(t+|c_i|u/|c_j|\). On this event, the orbit position at time \(t+u\) will be 
\[\displaystyle \bs A(t+u) = \cfrac{\bs \alpha e^{\bs S_j (|c_i|u)/|c_j|}}{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|) u}\bs e}.\] 
Correspondingly, on the event that at time \(t\) the level process of the fluid queue is \(X(t)=y_{k}\), the phase is \(\varphi(t)=j\) and there are no change of phase by time \(t+|c_i|u/|c_j|\), then \(X(t+u)=y_{k}+c_j|c_i|u/|c_j| = y_k+c_iu\). That is, \(X(t+u)\) is \(c_iu\) units from the left-hand edge of \(\calD_k\).

Now, observe 
\[\cfrac{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|)}}{\bs \alpha e^{\bs S_j (|c_i|u/|c_j|) u}\bs e} = \cfrac{\bs \alpha e^{\bs S_iu}}{\bs \alpha e^{\bs S_i u}\bs e} = \bs k(|c_i|u),\] 
which is exactly (\ref{eqn: orbit i}), the orbit position on the event \(E\). Hence, it is sensible to choose \(\bs D(i,j)=\bs I\).

@@

\subsection{On a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\)} 
Now suppose \(X(t)=y_k\), \(\varphi(t)=i\in\calS_+\), and the phase remains in state \(i\) until there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), \(u\in[0,\Delta/|c_i|)\). %We have reasoned, intuitively, that, for \(i\in\calS_+\) and for \(x=|c_i|u\in[0,\Delta)\), the orbit position \(\bs k(x)\) corresponds to the fluid level position \(X(t) = y_k+x\), that is \(x\) units greater than \(y_k\).  
As before, we need to find a matrix \(\bs D(i,j)\) to map the orbit position on the event \(E\) when there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), from \[\bs A(t+u^-) = \bs k(|c_i|u^-),\] 
to 
\[\bs A(t+u) = \cfrac{\bs k(u^-)\bs D(i,j)}{\bs k(u^-)\bs D(i,j)\bs e}=\cfrac{\bs\alpha e^{\bs S_iu} \bs D(i,j)}{\bs\alpha e^{\bs S_iu}\bs D(i,j)\bs e}\]
in such a way that the orbit position after the jump corresponds, in some way, to the fluid level being a distance of \(|c_i|u\) from \(y_k\) in phase \(j\).

@@

In the discussions above, we have relied on the relationship between the event \(E\) and the orbit position \(\displaystyle \bs k(|c_i|u)\). The relationship allows us to associate this orbit position with the level of the fluid queue. For phase-type distributions the vector \(\bs k(|c_i|u)\) is the vector of posterior probabilities \(\left[\mathbb P(J(u)=k \mid E)\right]_{k\in\{1,...,p\}}\). We can use this vector of posterior probabilities in the same way as we do for matrix-exponential distributions and QBD-RAPs. However, the orbit interpretation forgoes the presence of the phase process \(\{J(t)\}\). Here, we will use the phase process \(\{J(t)\}\) to derive a choice of \(\bs D(i,j)\).

Recall the position of the orbit process can be used to determine the residual time. This idea can be replicated when we are given the phase of the phase-type rather than the value of the orbit process. Given the phase of the phase-type distribution is \(k\in\{1,...,p\}\), the distribution of the residual time is 
\[\mathbb P(R_i\leq r \mid \mbox{phase}=k)=1-\bs e_k e^{\bs S_i r}\bs e.\]
Notice that this is independent of the time since the phase-type distribution was initialised. We call the time since the phase-type distribution was initialised the \emph{age}. As noted by \cite{hmp2017}, the distribution of the age given the phase of the phase-type is \(k\in\{1,...,p\}\) depends on the sampling scheme which determines the observation time. Here, on the event that at time \(t\) the phase of the QBD-RAP is \(i\), a change of phase occurs after an exponential amount of time with rate \(-T_{ii}\). Proposition 4.1, \cite{hmp2017} states that the distribution function of the age, given the phase of the phase-type is \(k\) and the process is observed after an exponential time with rate \(-T_{ii}\) is 
\[\mathbb P(\mbox{age}\leq u \mid \mbox{phase}=k )=1-\cfrac{\bs \alpha e^{(\bs S_i + T_{ii}\bs I)u}(-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e_k}{\bs \alpha (-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e_k}.\]
Let \(\widehat{ \bs S}_i(T_{ii}) = diag(\bs \nu)^{-1}\bs S_i'diag(\bs \nu)\), where \(\bs \nu = \bs\alpha(-(\bs S_i+ T_{ii}\bs I))^{-1}/(\bs\alpha(-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e)\). Algebraic manipulations show
\begin{align}
	1-\cfrac{\bs \alpha e^{(\bs S_i + T_{ii}\bs I)u}(-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e_k}{\bs \alpha (-(\bs S_i+ T_{ii}\bs I))^{-1}\bs e_k}
	= 1-\bs e_k' e^{(\widehat{\bs S}_i(T_{ii}) + T_{ii}\bs I)u} \bs e,\label{eqn: skgj}
\end{align}
which is of Phase-type.

Recall that in the matrix-exponential case we associated the value of the orbit process \(\bs k(|c_i|u)\) with being \(|c_i|u\) units to the right of \(y_{k}\). In contrast, for the phase-type case if we use the phase process \(\{J(t)\}\) to inform our modelling of the fluid queue (as opposed to the value of the orbit process), then we associate phase \(k\in\{1,...,p\}\) of \(\{J(t)\}\) with being a \emph{random} distance to the right of \(y_k\) where this distance has distribution 
\begin{align}
	1-\bs e_k' e^{(\widehat{\bs S}_i(T_{ii}) + T_{ii}\bs I)u} \bs e.\label{eqn: skgj2}
\end{align}
Therefore, on the event \(E\), upon a change from of phase of the QBD-RAP from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), it seems reasonable to want the distribution of time until the next event of the QBD-RAP to be
\begin{align}
	1-\bs e_k' e^{(\widehat{\bs S}_i(T_{ii}) + T_{ii}\bs I)|c_j|u/|c_i| + T_{jj}Iu} \bs e.\label{eqn: skgj3}
\end{align}
The factor \(|c_j|u/|c_i|\) arises as a conversion between the speed at which the fluid level moves in phase \(j\) compared to phase \(i\). The rate \(T_{jj}\) in the exponent is the rate at which the next change of phase occurs.

@@

For matrix exponential distributions we cannot rely on the phase process \(\{J(t)\}\) as we did in the phase-type case. Recall that we want to find a matrix \(\bs D(i,j)\) to map the orbit position on the event \(E\) when there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), from \(\bs A(t+u^-) = \bs k(|c_i|u)\), to 
\[\bs A(t+u) = \cfrac{\bs k(|c_i|u)\bs D(i,j)}{\bs k(|c_i|u)\bs D(i,j)\bs e}=\cfrac{\bs\alpha e^{\bs S_iu} \bs D(i,j)}{\bs\alpha e^{\bs S_iu}\bs D(i,j)\bs e}\]
in such a way that the orbit position after the jump corresponds, in some way, to the fluid level being a distance of \(|c_i|u\) from \(y_k\) in phase \(j\).

Given our interpretation of the orbit position, \(\bs k(x)\), a solution would be find a linear map which takes \(\bs k(x)\) and maps it directly to 
\[\bs k(\Delta-x) = \cfrac{\bs\alpha e^{\bs S(\Delta-x)} }{\bs\alpha e^{\bs S(\Delta-x) } \bs e}.\] However, we have been unsuccessful in finding such a mapping. The main hurdle being that the map must be linear. Instead, we approximate as follows.

Recall that we use \(R_i(u)\) to approximate the distance of the fluid level to the left of \(y_{k+1}\). Suppose we are given \(R_i(u)=r\) which corresponds, approximately, to the fluid level being \(|c_i|r\) units to the left of \(y_{k+1}\). The position of the orbit process which corresponds to a distance of \(|c_i|r\) to the left of \(y_{k+1}\) in phase \(j\) is \(\bs k(|c_i|r).\)
Hence, on \(R_i(u)=r\), the event \(E\) and on the event that a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) occurs at time \(t+u\), the orbit position should jump from \(\bs k(|c_i|u)\) to \(\bs k(|c_i|r)\). However, at time \(t+u\), \(R_i(u)\) is a random variable about the future of the process and therefore not known, so instead, we take the expected initial vector 
\[\mathbb E \left[ \bs k(|c_i|R_i(u))\right] = \int_{r=0}^\infty \cfrac{\bs\alpha e^{\bs S_i u } }{\bs\alpha e^{\bs S_iu } \bs e} e^{\bs S_ir } \bs s_i \bs k(|c_i|r)\wrt r = \cfrac{\bs\alpha e^{\bs S_i u } }{\bs\alpha e^{\bs S_iu } \bs e} \int_{r=0}^\infty e^{\bs S_ir } \bs s_i\cfrac{\bs\alpha e^{ \bs S_ir } }{\bs\alpha e^{ \bs S_ir } \bs e}\wrt r.\]
After a change of variables \(x=|c_i|r\) we get 
\[\cfrac{\bs\alpha e^{\bs S_i u } }{\bs\alpha e^{\bs S_iu } \bs e} \int_{x=0}^\infty e^{\bs S x }|c_i| \bs s \cfrac{\bs\alpha e^{ \bs Sx } }{\bs\alpha e^{ \bs Sx } \bs e}\wrt x/|c_i| = \bs A(t+u^-)\int_{x=0}^\infty e^{\bs S x }\bs s \cfrac{\bs\alpha e^{ \bs Sx } }{\bs\alpha e^{ \bs Sx } \bs e}\wrt x,\]
since at time \(t+u^-\), the orbit position is \(\bs A(t+u^-) = {\bs\alpha e^{\bs S_i u } }/{\bs\alpha e^{\bs S_iu } \bs e}\). Thus, we find 
\[\bs D(i,j)=\displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \bs k(x)\wrt x=:\bs D.\]

Observe that \(\bs D\bs e = \displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \bs k(x)\wrt x\bs e =\displaystyle \int_{x=0}^\infty e^{\bs S x }\bs s \wrt x=\bs e\), since \(\bs k(x)\bs e = 1\) for all \(x\geq 0\). Further, since \(\mathcal A\) is closed and convex \citep{MEinAP}, then \((\bs a \bs D,\bs S,\bs s)\) is a representation of a matrix exponential distribution for any \(\bs a\in\mathcal A\).

We pose the choice of the matrix \(\bs D\) as a modelling choice. Other choices are possible, for example, \(\displaystyle \bs D = \int_{x=0}^{\Delta}e^{\bs S x }\bs s \bs k(x)\wrt x\) is another sensible choice. It may also be possible to construct other matrices \(\bs D\), perhaps via geometric arguments.

@@

At time \(t\) an event occurs at rate 
\[\bs A(t) (\bs{S}_i - T_{ii}\bs{I}_p)\bs e = \bs A(t)\bs s_i - T_{ii}.\]
More precisely, an event corresponding to a change in phase for \({\phi}(t)\) occurs at rate \(-\bs A(t) T_{ii}\bs e = -T_{ii}\) and an event corresponding to a \emph{change of level} occurs at rate \(-\bs A(t) \bs{S}_i\bs e = \bs A(t) \bs s_i\). Later, we will make clear why we say that the latter event corresponds to a change of level. Upon an event occurring at time \(t\), with probability \({-T_{ii}}/{\left(-T_{ii} + \bs A({t^-})\bs s_i\right)}\) the event corresponds to a change of phase and with probability \({\bs A({t^-})\bs s_i}/{\left(-T_{ii} + \bs A({t^-})\bs s_i\right)}\) the event corresponds a change of level.
